---
title: "L'ORA - WG1 Working example Spillover (so.sc2)"
author: "Alfredo Acosta SVA" 
date: "`r Sys.Date()`" 
output: 
  html_document:
    toc: yes
    toc_float: true
    theme: cosmo
    fig_caption: yes
    number_sections: yes
    code_folding: show
  html_notebook: 
    toc: yes
---
# Working example SpillOver
-The code use several datasets EMPRES-I (FAO), ECDC Surveillance Atlas ECDC, Global Health observatory (WHO). Adagio.
-Please feel free to analyse the code and run from your terminal (there is a folder with the code and data on: DG1 Disease occurrence > datasources > Spill-over > sources > spillover.Rmd)

# Libraries
```{r}
# Spill over analysis
# Update: 17/01/2024
library(tidyverse)
library(lubridate)
library(stringr)
library(readr)
library(data.table)
library(plotly)
library(DT)
library(arsenal)
```
# Working directory
## Loading ECDC Surveillance atlas locally adquired cases
```{r}
setwd("C:/Users/alfredo.acosta/SVA/LiRA_consortium - Documents/WG1 Disease occurrence/datasources/Second-disease-group/spillover/locally adquired cases/")

# EGran
# soe <- read.csv(file="ECDC_surveillance_data_Echinococcosis.27.11.24.csv")
soe <- read.csv(file="ECDC_surveillance_data_Echinococcosis.csv")
# Lyme
soe1 <- read.csv(file="ECDC_surveillance_data_Lyme_Neuroborreliosis.csv")
# Qfever
soe2 <- read.csv(file="ECDC_surveillance_data_Q_fever.csv")
# TBE
soe3 <- read.csv(file="ECDC_surveillance_data_Tick-borne_encephalitis.csv")

# comparedf(soe1, soe2)
# comparedf(soe2, soe3)

# Influenza
# soe4 <- read.csv(file="ECDC_surveillance_data_Influenza.27.11.24.csv")
```


```{r}
soe <- rbind(soe, soe1, soe2, soe3)

table(soe$Indicator)

rm(soe1, soe2, soe3)

soe$NumValue <- gsub("-", "0", soe$NumValue)

soe1 <- soe %>% 
        filter(Unit == "N") %>% 
        filter(Indicator == "Reported cases") %>% 
        select(HealthTopic, RegionName, Time, NumValue)

colnames(soe1) <- c("disease", "country", "time", "cases")

soe2 <- soe %>% 
        filter(Unit == "%") %>% 
        filter(Indicator == "Completeness for travel association") %>% 
        select(HealthTopic, RegionName, Time, NumValue)

colnames(soe2) <- c("disease", "country", "time", "perc")
```
# Descriptive
```{r}
# Number of human cases
soe1 %>% 
  group_by(disease) %>%
  summarise(cases=sum(as.numeric(cases), na.rm = TRUE)) %>% 
  filter(cases != 0)

# Number of countries
soe1 %>% 
    group_by(disease) %>%
  summarise(countries=length(unique(country)))

table(soe$Time)
table(soe$Time, soe$HealthTopic)

```
# Loading Adagio database
```{r}
# Reading ADAGIO database
adg <- read.csv(file= "Outbreaks and cases with labels.csv")
str(adg)
```
# Loading Human population from Adagio(FAO)
```{r}
library(readxl)
adg_pop <- read_excel("FAO human population.xlsx")
```

# Matrix with human population over the years
```{r}
# Create matrix
db <- expand.grid(unique(adg_pop$Country), c(2003:2023), unique(soe1$disease))
colnames(db) <- c("country", "time", "disease")

# first index to  fill the human cases db & soe1
index <- match(paste(db$country, db$time, db$disease), paste(soe1$country, soe1$time, soe1$disease))
db$human_c_total <- as.numeric(soe1$cases[index])

# second index to pass the FAO Adagio human population
index <- match(paste(db$country, db$time), paste(adg_pop$Country, adg_pop$Year))

db$human_pop <- adg_pop$`Human population by country.Population in thousands`[index]

#Third index to pass the % of locally aquired cases
index <- match(paste(db$country, db$time, db$disease), paste(soe2$country, soe2$time, soe2$disease))
db$percentage_local <- soe2$perc[index]

db$human_pop <- db$human_pop*1000
```

# Locally adquired cases
```{r}
db$human_c_total <- as.numeric(as.character(db$human_c_total))
db$percentage_local <- as.numeric(as.character(db$percentage_local))
str(db)
db$percentage_local[db$percentage_local == 0] <- NA
table(is.na(db$percentage_local))

db$human_c_local <- round(ifelse(is.na(db$human_c_total), 0.000001, db$human_c_total) * 
             ifelse(is.na(db$percentage_local), 100, db$percentage_local + 0.0001) / 100, 2)

```

## Human cases / 100.000 habitants using FAO
```{r}
db$human_cases_hab <- db$human_c_local/db$human_pop*100000
db$human_cases_hab_total <- db$human_c_total/db$human_pop*100000
```


# Locally adquired cases and total
```{r}
summary(db$human_cases_hab) #locally
summary(db$human_cases_hab_total) # all

summary(db$percentage_local) #percentage of local

```


# Reading the UN regions
```{r}
unregions <- read.csv("all.csv")
unregions$name[unregions$name == "United States of America"] <- "United States"
unregions$name[unregions$name == "United Kingdom of Great Britain and Northern Ireland"] <- "United Kingdom"

str(unregions)
str(db)
db$country <- as.character(db$country)
db <- data.frame(db)
db$unr <- unregions$sub.region[match(db$country, unregions$name)]

```



# General view on rate x 100.000 hab locally adquired
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, group=country))+
  geom_point()

db %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, group=country))+
  geom_point()+
  scale_y_log10()+
  facet_wrap(vars(disease), ncol = 1)
```
# General view on rate x 100.000 hab total cases
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab_total, group=country))+
  geom_point()

db %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab_total, group=country))+
  geom_point()+
  scale_y_log10()+
  facet_wrap(vars(disease), ncol = 1)
```


```{r}
# General plots by country
# List of diseases
diseases <- unique(db$disease)

# Function to generate plots for a given disease
generate_plots <- function(disease_name) {
  # Filter data for the specific disease
  disease_data <- db %>% 
    filter(time > 2003, disease == disease_name, !is.na(human_cases_hab))
  
  # Check if disease_data is empty
  if (nrow(disease_data) == 0) {
    return(list(plot1 = NULL, plot2 = NULL, plot3 = NULL))
  }
  
  # Plot 1: Boxplot per country
  plot1 <- disease_data %>%
    ggplot(aes(human_cases_hab, color = country)) +
    geom_boxplot() +
    scale_x_log10() +
    facet_wrap(vars(country), ncol = 1) +
    theme_minimal() +
    theme(legend.position = "none") +
    xlab(paste(disease_name, "human cases / 100.000 habitants")) +
    ylab(NULL) +
    guides(y = "none")
  
  # Plot 2: Boxplot grouped by year
  plot2 <- disease_data %>%
    ggplot(aes(human_cases_hab, time, group = time)) +
    geom_boxplot() +
    scale_x_log10() +
    theme_minimal() +
    theme(legend.position = "none") +
    xlab(paste(disease_name, "human cases / 100.000 habitants (boxplot grouped by year)")) +
    ylab(NULL)
  
  # Plot 3: Time-series plot
  plot3 <- disease_data %>%
    ggplot(aes(time, human_cases_hab, color = country)) +
    geom_point() +
    geom_line(size = 1) +
    scale_y_log10() +
    facet_wrap(vars(country), ncol = 3) +
    theme_minimal() +
    theme(legend.position = "none") +
    xlab(paste(disease_name, "human cases / 100.000 habitants (by year-country)")) +
    ylab(NULL)
  
  # Return a list of plots
  list(plot1 = plot1, plot2 = plot2, plot3 = plot3)
}

# Generate all plots for all diseases
all_plots <- map(diseases, ~ generate_plots(.x))

names(all_plots) <- diseases

all_plots

# Access plots for specific diseases
# Qfever_plots <- all_plots[["Q fever"]]
# EGran_plots <- all_plots[["Echinococcosis"]]
# Lyme_plots <- all_plots[["Lyme Neuroborreliosis"]]
# TBE_plots <- all_plots[["Tick-borne encephalitis"]]

# Display specific plots
# Qfever_plots$plot1
# Qfever_plots$plot2
# Qfever_plots$plot3

# EGran_plots$plot1
# Lyme_plots$plot2
# TBE_plots$plot3
```

# General view of diseases spread
```{r}
db %>% 
  filter(time > 2003) %>% 
  # filter(disease == "Rabies") %>% 
  # filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time, color=disease))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(disease), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("human cases/100.000 habitants (log scaled) country rates")

db %>% 
  filter(time > 2003 & !is.na(human_cases_hab | !is.na(disease))) %>% 
  ggplot(aes(human_cases_hab, unr, group=unr, color=disease))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(disease), scales="free", ncol = 1)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("human cases/100.000 habitants (log scaled) country rates ")
```

# Using UN regions
```{r}
diseases <- unique(db$disease)

generate_plots_unr <- function(disease_name) {
  
  disease_data <- db %>% 
    filter(time > 2003, disease == disease_name, !is.na(human_cases_hab))

  #boxplot by unr 
p1 <- disease_data %>% 
  ggplot(aes(human_cases_hab, color=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(unr), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab(paste(disease_name, "human cases / 100.000 habitants"))+
  ylab(NULL) + 
  guides(y="none")

# boxplot by year
p2 <- disease_data %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab(paste(disease_name, "human cases / 100.000 habitants"))+
  ylab(NULL)

# Time series plot
p3 <- disease_data %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(unr), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab(paste(disease_name, "human cases / 100.000 habitants (by year-unr)"))

# Return a list or plots
  list(p1 = p1, p2 = p2, p3 = p3)
}

# Generate all plots for all diseases
all_plots2 <- map(diseases, generate_plots_unr)

all_plots2

# Access plots for a specific disease, e.g., "Q fever"

# Qfever_plots_unr<- all_plots2[[which(diseases == "Q fever")]]
# EGran_plots_unr <- all_plots2[[which(diseases == "Echinococcosis")]]
# Lyme_plots_unr <- all_plots2[[which(diseases == "Lyme Neuroborreliosis")]]
# TBE_plots_unr <- all_plots2[[which(diseases == "Tick-borne encephalitis")]]


```

# Calculating cut points and levels by disease and UNregions
```{r}

UNRegion_db_unr <- db %>% 
  group_by(disease, unr) %>% 
  filter(time > 2003) %>% 
  filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>%
  summarise(human_cases_hab_unr=(sum(human_c_local)/sum(human_pop)*100000)) %>% 
  mutate(name=cut(human_cases_hab_unr, breaks = 3, labels = c("low","medium","high"))) %>% 
  mutate(range=cut(human_cases_hab_unr, breaks = 3))
UNRegion_db_unr


UNRegion_db <- db %>% 
  group_by(disease) %>% 
  filter(time > 2003) %>% 
  filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>%
  summarise(human_cases_hab_unr=(sum(human_cases_hab)/sum(human_pop)*100000)) %>% 
  mutate(name=cut(human_cases_hab_unr, breaks = 3, labels = c("low","medium","high"))) %>% 
  mutate(range=cut(human_cases_hab_unr, breaks = 3))
UNRegion_db

```

# General cut points by disease
```{r}

```

# Lira table median 95 CI
```{r}
lira_table_sp <- db %>%
  filter(time > 2003) %>%
  filter(human_cases_hab > 0) %>% 
  filter(!is.na(human_cases_hab)) %>%
  group_by(disease, UNregions=unr) %>%
  summarise(
    human_cases=sum(human_c_local),
    human_pop=median(human_pop, na.rm = TRUE),
    human_rate_median=round(quantile(human_cases_hab, probs = 0.5, na.rm = TRUE), 3),
    CI05 = round(quantile(human_cases_hab, probs = 0.025, na.rm = TRUE), 3),
    CI95 = round(quantile(human_cases_hab, probs = 0.975, na.rm = TRUE) ,3)) %>%
  mutate(level=cut(human_rate_median, breaks = 3, labels = c("low","medium","high"))) %>% 
  arrange(disease, desc(human_rate_median), UNregions)

datatable(lira_table_sp)

# lira_table_sp
# write.csv(lira_table_sp, file="lira_table_sp_sc2.csv")
```

# database for adagio use
```{r}
db_adagio <- db %>% 
  filter(!human_c_local ==0)   

# write.csv(db_adagio, file = "ECDC_adagio.csv")

```


```{r}
lira_table_sp_total <- db %>%
  filter(time > 2003) %>%
  filter(human_cases_hab > 0) %>%
  filter(!is.na(human_cases_hab_total)) %>%
  group_by(disease, UNregions = unr) %>%
  summarise(
    human_cases = sum(human_c_total, na.rm = TRUE),
    human_c_total_sum = sum(human_c_total, na.rm = TRUE),
    human_c_local_sum = sum(human_c_local, na.rm = TRUE),
    diference = round((human_c_total_sum - human_c_local_sum) / human_c_total_sum * 100,2),
    human_pop = median(human_pop, na.rm = TRUE),
    human_rate_median = round(quantile(human_cases_hab_total, probs = 0.5, na.rm = TRUE), 3),
    CI05 = round(quantile(human_cases_hab_total, probs = 0.025, na.rm = TRUE), 3),
    CI95 = round(quantile(human_cases_hab_total, probs = 0.975, na.rm = TRUE), 3),
    .groups = "keep"
  ) %>%
  mutate(level = cut(human_rate_median, breaks = 3, labels = c("low", "medium", "high"))) %>%
  arrange(disease, desc(human_rate_median), UNregions)

datatable(lira_table_sp_total)

```


#Graph 
```{r}
library(ggplot2)

# Visualization
ggplot(lira_table_sp, aes(x = reorder(UNregions, -human_rate_median), y = human_rate_median, color = disease)) +
  geom_point(size = 3, position = position_dodge(width = 0.1)) +  # Points for median
  geom_errorbar(aes(ymin = CI05, ymax = CI95), width = 0.2, position = position_dodge(width = 0.5)) +  # Error bars for CI
  facet_wrap(vars(disease), scales = "free_y") +  # Separate panels for each disease
  coord_flip() +  # Flip axes for better readability
  scale_y_log10()+
  theme_minimal() +
  labs(
    title = "Median Human Cases and 95% CI by Disease and UN Region",
    x = "UN Region",
    y = "Locally acquired Human Cases / 100,000 habitants",
    color = "Disease"
  ) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplot(lira_table_sp_total, aes(x = reorder(UNregions, -human_rate_median), y = human_rate_median, color = disease)) +
  geom_point(size = 3, position = position_dodge(width = 0.1)) +  # Points for median
  geom_errorbar(aes(ymin = CI05, ymax = CI95), width = 0.2, position = position_dodge(width = 0.5)) +  # Error bars for CI
  facet_wrap(vars(disease), scales = "free_y") +  # Separate panels for each disease
  coord_flip() +  # Flip axes for better readability
  scale_y_log10()+
  theme_minimal() +
  labs(
    title = "Median Human Cases and 95% CI by Disease and UN Region",
    x = "UN Region",
    y = "Total reported Human Cases / 100,000 habitants",
    color = "Disease"
  ) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

# Graph 
```{r}
ggplot(lira_table_sp, aes(x = level, y = human_rate_median, fill = level)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +  # Boxplot for distribution
  geom_jitter(aes(color = level), width = 0.2, size = 2, alpha = 0.6) +  # Add points for individual medians
  facet_wrap(vars(disease), scales = "free_y") +  # Separate panels for each disease
  theme_minimal() +
  labs(
    title = "Distribution of Human Rate Median by Level (High, Medium, Low)",
    x = "Level",
    y = "Human Cases / 100,000 habitants",
    fill = "Level",
    color = "Level"
  ) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines")
  )

```

# Credits Acosta, Alfredo PhD<sup>1</sup>.
**SVA<sup>1</sup>**: SVA <http://www.sva.se/>. 